---
title: "RNA-Seq Analysis of NMR and MM cell lines"
author: "Karsten Bach"
date: '`r Sys.Date()`'
output:
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: journal
    highlight: tango
    code_folding: hide
---
***


# Load Data
## Count matrices
- The two datasets are read in independently
- STAR adds statistics such as unmapped, ambiguous reads etc. that I extract first
```{r, message=FALSE}
library(edgeR)
library(viridis)
library(cowplot)
library(ggplot2)
library(biomaRt)
library(dplyr)
library(pheatmap)
theme_set(theme_cowplot())

# Read in Counts
# Mouse
mm.counts <- read.csv("../data/counts/MM_CountMatrix.csv",row.names=1)
pDat <- read.csv("../data/meta/SampleInfo.csv",stringsAsFactors=FALSE)
pDat_mm <- pDat[pDat$Species=="Mouse",]
pDat_nmr <- read.csv("../data/meta/NMR_ColData.csv",stringsAsFactor=FALSE)[,-1]
# NMR
nmr.counts <- read.csv("../data/counts/NMR_Counts_Clean.csv",row.names=1)

# Move stats to PD
stopifnot(identical(c(colnames(mm.counts)),pDat_mm$Sample))
pDat_mm$Unmapped <- t(mm.counts)[,1]
pDat_mm$MultiMap <- t(mm.counts)[,2]
pDat_mm$NoFeat <- t(mm.counts)[,3]
pDat_mm$Ambiguous <- t(mm.counts)[,4]
mm.counts <- mm.counts[-c(1:4),]
pDat_mm$TotalReads <- colSums(mm.counts)
pDat <- rbind(pDat_mm,pDat_nmr)
```


## Combining the NMR and MM data
- I use a published list of orthologs mapping roughly ~16k genes from mouse to nmr (**Hilton et al.**)
- I then concatenate the two gene expression matrices and use the mouse gene symbols as identifiers
```{r, message=FALSE}

# Subset both datasets to orthologous genes
orthologs <- read.csv("../data/meta/media-28.csv",stringsAsFactor=FALSE)
colnames(orthologs)[1] <- "entrezgene"
orthologs <- orthologs[!orthologs$entrezgene %in% setdiff(orthologs$entrezgene,rownames(nmr.counts)),] # They have about a 100 genes in the list that are not annotated
orthologs <- orthologs[!orthologs$mouse.gene_id %in% setdiff(orthologs$mouse.gene_id,rownames(mm.counts)),] # They have about a 5 genes in the list that are not annotated
# Add in the transgenes to the orthologs list
nmr.genes <- orthologs$entrezgene
mm.genes <- orthologs$mouse.gene_id
# Subset to ortholog set
nmr.counts <- nmr.counts[nmr.genes,]
mm.counts <- mm.counts[mm.genes,]
# Set same names
rownames(nmr.counts) <- rownames(mm.counts)
counts <- cbind(mm.counts,nmr.counts)

# Add Gene Annotation
# Using the orthologs df now
orthologs$uniq <- scater::uniquifyFeatureNames(orthologs$mouse.gene_id, orthologs$mouse.symbol)
rownames(counts)[rownames(counts) %in% orthologs$mouse.gene_id] <- orthologs$uniq

# gene.data$uniq <- scater::uniquifyFeatureNames(gene.data[,1],gene.data[,2])
# gene.data <- gene.data[rownames(counts)[-((nrow(counts)-5):nrow(counts))],]
# rownames(counts) <- c(gene.data$uniq,rownames(counts)[(nrow(counts)-5):nrow(counts)])


rownames(pDat) <- pDat$Sample
stopifnot(identical(rownames(pDat), colnames(counts)))
```

# Sanity checks
## QC stats
- Here I'm just checking that there is no massive differences in any of the statistic for any particular sampel
```{r, message=FALSE, fig.width=9}
p0 <- ggplot(pDat, aes(x=Sample,y=Unmapped)) +
    geom_bar(stat="identity") +
    facet_wrap(~Species,scale="free_x")
p1 <- ggplot(pDat, aes(x=Sample,y=MultiMap)) +
    geom_bar(stat="identity") +
    facet_wrap(~Species,scale="free_x")
p2 <- ggplot(pDat, aes(x=Sample,y=Ambiguous)) +
    geom_bar(stat="identity") +
    facet_wrap(~Species,scale="free_x")
p3 <- ggplot(pDat, aes(x=Sample,y=NoFeat)) +
    geom_bar(stat="identity") +
    facet_wrap(~Species,scale="free_x")
p4 <- ggplot(pDat, aes(x=Sample,y=TotalReads)) +
    geom_bar(stat="identity") +
    facet_wrap(~Species,scale="free_x")
plot_grid(p0,p1,p2,p3,p4)
```

```{r, message=FALSE}
# Set up DGEList
keep <- rowSums(counts)>0
counts <- counts[keep,]
y <- DGEList(counts=counts,samples=pDat)
# Normalize
y <- calcNormFactors(y)

# Get Log-Cpm
logcpm <- cpm(y,log=TRUE)
```

# Principal Component Analysis
- PCA on normalized log transformed counts of all orthologs
```{r, message=FALSE, fig.width=7}
# Compute PCA without transgenes
pcs <- prcomp(t(logcpm))
logcpm.pc <- pcs$x
vars <- apply(logcpm.pc,2,var)
rel.vars <- round(vars/sum(vars)*100,1)

pDat <- y$samples
pDat$PC1 <- logcpm.pc[,1]
pDat$PC2 <- logcpm.pc[,2]
pDat$PC3 <- logcpm.pc[,3]

out <- group_by(pDat, Species, Transformation, Source) %>%
    summarize(mPC2=mean(PC2))
write.csv(out,"pc2s.csv")

pDat$Transformation <- factor(pDat$Transformation, levels=c("Untransformed","SV40LT","SV40LTRasHadi","SV40LTRasTian"))
# Plot
ggplot(pDat, aes(x=PC1, y=PC2, color=Transformation, size=Species, shape=Source)) +
    geom_point() +
    xlab(paste0("PC1(",rel.vars["PC1"],"%)")) +
    ylab(paste0("PC2(",rel.vars["PC2"],"%)")) +
    scale_color_manual(values=c("black","#4DBBD5FF","red","#3C5488FF")) +
    coord_fixed(1) +
    scale_size_manual(values=c(2.5,4))
ggsave("PCA.svg")
```

# Loadings
- We can now look at the loadings of PC2 to see what processes are involved in separating the cells along an axis of transformation
```{r, message=FALSE, fig.width=7}
ld.pc2 <- pcs$rotation[,2]
# plot(ld.pc2[order(ld.pc2,decreasing=TRUE)],ylab="Loadings of PC2")
# abline(h=0.02,col="red",lty="dashed")
# abline(h=-0.02,col="red",lty="dashed")
ld.pc2.ord <- ld.pc2[order(ld.pc2,decreasing=TRUE)]
fp <- data.frame("Gene"=names(ld.pc2.ord),
		 "Loading"=ld.pc2.ord)
fp$Index <- c(1:nrow(fp))
library(ggrepel)
ggplot(fp, aes(x=Index, y=Loading)) +
    geom_point(color="grey80") +
    geom_point(data=fp[abs(fp$Loading)>0.02,],color="black") +
    geom_text_repel(data=fp[c(1:10,(nrow(fp)-10):nrow(fp)),], aes(label=Gene)) +
    geom_hline(yintercept=0.02,color="red",lty="dashed") +
    geom_hline(yintercept=-0.02,color="red",lty="dashed") +
    ylab("PC2 Loading")
ggsave("Loadings.svg")
```

## GO - Analysis
```{r, message=FALSE, fig.width=9}
ps <- ld.pc2[ld.pc2 > 0.02]
ps <- ps[order(ps,decreasing=TRUE)]
neg <- ld.pc2[ld.pc2 < -0.02]
neg <- neg[order(neg)]

topLoads <- list("Positive Loading"=ps, "Negative Loading"=neg)# topGO
univrs <- names(ld.pc2)
out <- list()

for (i in seq_along(topLoads)) {
    gens <- names(topLoads[[i]])
    subst <- names(topLoads)[[i]]
    # Gene universe
    alG <- factor(as.numeric(univrs %in% gens))
    names(alG) <- univrs

    # ---- GOanalysis ----

    library(topGO)
    #prepare Data for topGO
    GO.data <- new("topGOdata", description="Lib GO",ontology="BP", allGenes=alG, 
		   annot=annFUN.org, mapping="org.Mm.eg.db",
		   nodeSize=20, ID="symbol")
    result.classic <- runTest(GO.data, statistic="fisher")
    res <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=300)

    res$Fisher.classic <- as.numeric(res$Fisher.classic)
    res$Term <- factor(res$Term, levels=res$Term[order(res$Fisher.classic,decreasing=TRUE)])
    out[[subst]] <- res[1:5,] %>%  # this only works because they are ordered by fisher classic
			mutate(hitsPerc=Significant*100/length(gens)) %>% 
			ggplot(aes(y=hitsPerc, 
				   x=Term, 
				   fill=Fisher.classic
				   )) +
			    geom_col() +
			    expand_limits(x=0) +
			    labs(y="Hits (%)", x="GO term", colour="p value", size="Count") +
			    scale_fill_viridis(begin=1,end=0) +
			    coord_flip()
}
out[[1]] %+% ggtitle(names(out)[1])
ggsave("Positive.svg",width=12)
out[[2]] %+% ggtitle(names(out)[2])
ggsave("Negative.svg",width=12)
```
